---
title: "CleanLASA introduction"
author: "B.Hoogerheide"
output: 
  html_document: 
    toc: false
    keep_md: true
cache: true
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(CleanLASA)
```

## Overview

`CleanLASA` is designed to make LASA (Longitudinal Aging Study Amsterdam) SPSS data easier to find, import, document, and prepare for analysis in R. The package connects three parts of the LASA workflow that otherwise require substantial manual work:

1.  finding the relevant LASA topic, file code, measurement waves, and variable documentation;
2.  importing LASA `.sav` files while applying the correct wave-specific variable and value labels; and
3.  auditing, refreshing, or manually correcting the metadata used for labelling.

The central import function is `read_lasa_sav()`. It identifies the LASA wave and file code from the file name and then applies metadata from the package's normalized label database. The same labelling engine is available separately through `apply_lasa_labels()` when data have already been imported.

A typical workflow is therefore:

``` text
find topic/file code -> inspect LASA documentation -> import .sav file
        -> apply/reshape labels -> inspect matching audit -> analyse data
```

The examples involving LASA data files or the live LASA website are not run when this vignette is built, because LASA data are not distributed with the package and website access may not be available during package installation.

## 1. Find topics, file codes, and waves

Use `lasa_topics()` to search the LASA topic overview. With no arguments it returns the complete topic index; searches can instead be restricted by topic, theme, or subtheme.

```{r topics, eval=FALSE}
# Browse the complete LASA topic index
lasa_topics()

# Search by topic name; partial and small fuzzy matches are supported
lasa_topics(topic = "physical act")

# Search a theme or subtheme
lasa_topics(theme = "cognitive")
lasa_topics(theme = "memory")
```

The returned data frame contains one row per topic/file-code combination, with columns for `theme`, `subtheme`, `topic`, `filecode`, `waves`, and `has_varinfo`. This makes it possible to identify not only the relevant file code, but also the waves in which that topic is available and whether LASA links a variable-information PDF for it.

The topic table is retrieved from the LASA website and cached both for the current R session and on disk. Set `refresh = TRUE` when you explicitly want to rebuild the cache from the current live topic table:

```{r topics-refresh, eval=FALSE}
lasa_topics(topic = "physical activity", refresh = TRUE)
```

## 2. Open the corresponding variable-information PDF

After identifying a topic or file code, `lasa_var_info()` resolves the variable-information PDF linked by LASA. It accepts common variations of a file code as well as topic names.

```{r var-info, eval=FALSE}
# By file code
lasa_var_info("046")
lasa_var_info("LASA046")

# By topic name
lasa_var_info("Physical activity")
lasa_var_info("physical act")

# Retrieve the URL without opening the PDF
url <- lasa_var_info("046", open = FALSE)
```

By default, an interactive R session opens the PDF using the requested viewer. No LASA variable-information PDFs are bundled with `CleanLASA`; this function resolves the document from the LASA topic index.

## 3. Import and label a LASA SPSS file

`read_lasa_sav()` is the main entry point for analysis data. It reads the file with `haven::read_sav()`, derives the LASA wave and file code from the file name, retrieves the corresponding metadata from `lasa_label_db()`, and applies variable and value labels.

```{r import, eval=FALSE}
# Regular LASA wave: LASA + wave + file code
physical_activity_b <- read_lasa_sav("data/LASAB046.sav")

# Information stored in a Z file
sex <- read_lasa_sav("data/LASAZ004.sav")

# Replenishment/migrant-cohort waves use LAS2B/LAS3B/LAS4B/LASMB
physical_activity_3b <- read_lasa_sav("data/LAS3B046.sav")
```

The filename parser is case-insensitive. Regular single-letter waves follow `LASA[wave][filecode].sav`; waves `2B`, `3B`, `4B`, and `MB` follow `LAS[wave][filecode].sav`.

By default, `user_na = TRUE` is passed to `haven::read_sav()` so SPSS user-defined missing codes remain available while labels and optional conversions are applied. Additional `haven::read_sav()` arguments can be supplied through `read_sav_args`:

```{r import-extra, eval=FALSE}
data <- read_lasa_sav(
  "data/LASAB046.sav",
  read_sav_args = list(encoding = "UTF-8")
)
```

Do not include `file` or `user_na` inside `read_sav_args`; those are controlled by `path` and `user_na` directly.

After import, the returned object carries LASA provenance attributes for the wave, file code, and source filename. The labelling audit is stored separately as the `label_report` attribute.

## 4. Choose how labelled variables are represented

The default behavior preserves the imported values while attaching the variable label (`attr(x, "label")`) and, where documented, the SPSS value-label map (`attr(x, "labels")`). Several arguments can additionally reshape the result for analysis.

```{r raw_data}
data <- haven::read_sav("../data/LASAB046.sav")
```

```{r collapse=TRUE}
summary(data$blphya01)

attr(data$blphya01, "labels")
```

### Convert labelled categories to factors

Use `to_factor = TRUE` when you want documented value labels represented as R factor levels:

```{r factors}
data <- read_lasa_sav(
  "../data/LASAB046.sav",
  to_factor = TRUE
)
summary(data$blphya01)

attr(data$blphya01, "original_labels")
```

Observed values that do not have a documented codebook label are retained as their numeric code (represented as text) rather than silently converted to `NA`.

### Restore count/continuous variables to plain numeric

The label database classifies variables whose codebook contains no categorical codes (or only negative missing-reason codes) as numeric. With `to_numeric = TRUE`, those variables are restored to ordinary numeric vectors and negative observed codes are converted to `NA`:

```{r numeric}
data <- read_lasa_sav(
  "../data/LASAB046.sav",
  to_numeric = TRUE
)
summary(data$blphya01)

attr(data$blphya01, "original_labels")
```


`to_factor` and `to_numeric` can be used together. In that workflow, numeric-classified variables are restored to numeric while other variables with value-label maps can be represented as factors.

### Standardize variable names and separate the wave code

LASA variable names often contain a wave-specific prefix. For example, a variable may be documented as `blphya01` in wave B but have the canonical name `lphya01` in the label database.

`standardize_names = TRUE` renames matched variables to their canonical lowercase names, standardizes the respondent-number column to `respnr`, and also implies `split_wavecode = TRUE`. A new `LASA_wave` column is inserted immediately after `respnr` when that identifier is present.

```{r standardize, collapse=TRUE}
data <- read_lasa_sav(
  "../data/LASAB046.sav",
  standardize_names = TRUE
)

names(data)[1:4]
unique(data$LASA_wave)
```

`split_wavecode = TRUE` can also be used on its own. It renames matched LASA variables to their canonical names and inserts `LASA_wave`, but does not by itself force a differently-capitalized respondent-number column to be renamed to `respnr`.

```{r split-wave, eval=TRUE}
data <- read_lasa_sav(
  "../data/LASAB046.sav",
  split_wavecode = TRUE
)

names(data)[1:4]
unique(data$LASA_wave)
```

### Correct a known source-column name

If a source file contains a known typo or nonstandard column name, use `name_corrections` to map a documented wave-specific or canonical variable name to the actual source column:

```{r name-corrections, eval=FALSE}
data <- read_lasa_sav(
  "../data/LASAB046.sav",
  name_corrections = c(lphya08 = "BLPYA08")
)
```

This override is used before the automatic exact/case-insensitive matching steps.

### Preserve the original coding for reference

For matched variables, `CleanLASA` stores the original imported values in the `original_values` attribute. When a documented value-label map exists, it is also retained in `original_labels`. These reference attributes remain available after factor/numeric reshaping, which makes it possible to compare the transformed R representation with the original SPSS coding.

```{r original-coding, eval=FALSE}
attr(data$lphya01, "original_values")
attr(data$lphya01, "original_labels")
```

## 5. Audit how variables were matched

Every call to `read_lasa_sav()` or `apply_lasa_labels()` attaches a variable matching audit. Retrieve it with `lasa_label_report()`:

```{r audit, eval=FALSE}
report <- lasa_label_report(data)
report
```

By default, this returns the **full** audit, including successful matches and documented variables that were not found in the data. The `method` column records how a match was made, for example `exact`, `case-insensitive exact`, `exact canonical`, `manual correction`, or `not found`.

To focus only on entries that may require attention, use:

```{r audit-problems, eval=FALSE}
lasa_label_report(data, problems_only = TRUE)
```

This is useful before analysis when you want to verify that expected variables were matched correctly. Columns in your data that are not documented for the selected file code/wave (for example newly created analysis variables) are not modified by the label engine.

## 6. Apply LASA labels to an existing data frame

`apply_lasa_labels()` exposes the same database-driven label engine without reading a file. This is useful if you imported a `.sav` file yourself or need to re-apply metadata after transformations.

```{r apply-labels, eval=FALSE}
raw <- haven::read_sav("data/LASAC046.sav", user_na = TRUE)

data <- apply_lasa_labels(
  raw,
  filecode = "046",
  wave = "C",
  to_factor = TRUE,
  to_numeric = TRUE,
  standardize_names = TRUE
)
```

When `data` already carries the `LASA_file_code` and `LASA_wave` provenance attributes created by `read_lasa_sav()` or a previous `apply_lasa_labels()` call, those arguments can usually be omitted:

```{r reapply-labels, eval=FALSE}
data <- read_lasa_sav("../data/LASAB046.sav")

# ...perform transformations that may remove column attributes...

data <- apply_lasa_labels(data)
```

For the wave, `apply_lasa_labels()` can also use a single-valued `LASA_wave` column. Supplying `filecode` and `wave` explicitly is the most robust option for data frames that did not originate from `read_lasa_sav()`.

## 7. Inspect the label database

`lasa_label_db()` returns the metadata database currently in effect. It is a list with four main components:

- `variables`: one row per file-code/wave/variable combination, including the wave-specific name, canonical name, wave-specific variable label, cross-wave-consistent ("harmonized") variable label, and variable type;
- `value_labels`: the long-format value/code label definitions, as documented for that specific wave;
- `value_labels_harmonized`: the same variable's value labels standardized across every wave that documents it (no `wave` column -- it applies regardless of wave), for combining data across waves that coded the same concept slightly differently; and
- `manual_overrides`: separately stored hand-authored corrections for variables and value labels.

```{r inspect-db, eval=FALSE}
db <- lasa_label_db()

# Which wave-B variables are currently known for file code 046?
subset(db$variables, filecode == "046" & wave == "B")

# Inspect the corresponding value-label rows
subset(db$value_labels, filecode == "046" & wave == "B")

# The cross-wave-standardized value labels for the same variable
subset(db$value_labels_harmonized, filecode == "046" & canonical_name == "lphya01")
```

The package ships with a bundled snapshot, transcribed from LASA's own variable-information documentation; coverage grows and gets corrected with package updates. If you record a manual correction (below), `CleanLASA` writes a full updated snapshot to the package-specific user data directory and uses that user-local database in preference to the bundled copy on subsequent calls. This means corrections persist across R sessions without modifying the installed package library.

## 8. Record a manual metadata correction

`manual_update_lasa_labels()` is an escape hatch for a source-document error or a variable/value label that needs a deliberate local correction. The correction is stored in a separate override layer, so a later package update does not silently remove it -- it only replaces the bundled base layer the correction is composed on top of.

Merge one corrected value label into the existing set:
With `replace_val_labels=FALSE` (the default), supplied value labels are appended to the existing documentation.

```{r manual-update-merge, eval=FALSE}
manual_update_lasa_labels(
  filecode = "046",
  wave = "B",
  variable = "lphya01",
  val_labels = c(`-5` = "NA, wrong, skip")
)
```

Or replace the complete value-label set for the variable across all waves in which that canonical variable is represented:
With `replace_val_labels=TRUE`, the value labels in the label database are overridden with the supplied values, thus ignoring the original value labels. 

```{r manual-update-replace, eval=FALSE}
manual_update_lasa_labels(
  filecode = "046",
  wave = "all",
  variable = "lphya01",
  val_labels = c(`-5` = "NA, wrong, skip"),
  replace_val_labels = TRUE
)
```

The same function can replace a variable label through `var_label` to replace the existing variable labels.

```{r manual-update-varlabel, eval=FALSE}
manual_update_lasa_labels(
  filecode = "046",
  wave = "B",
  variable = "lphya01",
  var_label = "New label"
)
```

Both functionalities can also be combines and/or applied across multiple waves at once:
This updates the local database, and can subsequently be applied to the loaded dataframes with `apply_lasa_labels()`

```{r manual-update-full, eval=FALSE}
manual_update_lasa_labels(
  filecode = "046",
  wave = "all", # update all waves. Currently B-K
  variable = "lphya01",
  var_label = "New label",
  val_labels = c(`-5` = "NA, wrong, skip")
)

manual_update_lasa_labels(
  filecode = "046",
  wave = c("B", "C", "D"), # only update specified waves.  
  variable = "lphya01",
  var_label = "New label",
  val_labels = c(`-5` = "NA, wrong, skip")
)
```

## 9. A complete analysis-oriented workflow

For most analyses, the package can be used with a small number of steps:

```{r complete-workflow, eval=FALSE}
library(CleanLASA)

# 1. Find the relevant LASA topic and file code
lasa_topics(topic = "physical activity")

# 2. Inspect the official variable-information document if needed
lasa_var_info("046")

# 3. Import the LASA file, convert analysis variables, and standardize names
data <- read_lasa_sav(
  "../data/LASAB046.sav",
  to_factor = TRUE,
  to_numeric = TRUE,
  standardize_names = TRUE
)

# 4. Check only matching problems before analysis
problems <- lasa_label_report(data, problems_only = TRUE)
problems
```

If the required file code is not yet covered by the active label database, `read_lasa_sav()` does not dispatch to a file-specific labelling function or invent metadata. Add a deliberate correction with `manual_update_lasa_labels()`, or wait for a package update that adds coverage, and import/apply the labels again.

## Function summary

| Function | Main purpose |
|--------------------|----------------------------------------------------|
| `lasa_topics()` | Search LASA topics, themes/subthemes, file codes, and available waves |
| `lasa_var_info()` | Resolve and open the LASA variable-information PDF for a file code or topic |
| `read_lasa_sav()` | Read a LASA `.sav` file, identify wave/file code, and apply database-backed labels |
| `apply_lasa_labels()` | Apply the same LASA labelling engine to an existing data frame |
| `lasa_label_report()` | Inspect the full matching audit or only entries requiring attention |
| `lasa_label_db()` | Inspect the active bundled + user-local label metadata database |
| `manual_update_lasa_labels()` | Add or correct metadata in a persistent manual override layer |

Together, these functions make the LASA documentation part of the R data workflow: topics and source documents remain discoverable, imported variables retain their documented meaning, transformations remain auditable, and the metadata layer can be refreshed without maintaining separate file-specific labelling functions.

```         
```
